Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2020-07-09 ms/ms.tex
OU comments
revised for submission

Here, we estimate the joint posterior distribution of the \emph{effective} start date of the NPIs $\tau$ and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$. 
We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
This is done under the simplifying assumption that all interventions start at a specific date, despite the reality that the durations between the first and last NPIs were between 4 and 12 days (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant).
All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~24, and these predictions were then compared to the real case data. 
The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark. Notably, the data for Sweden and Denmark does not have a peak during the evaluated dates, possibly leading to wide 95\% confidence intervals on $\hat{tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).

We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$; (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
The exceptions are Switzerland (see below), and Norway.
The latter also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
In the following, we describe our findings in more detail.

%%% Late
\paragraph*{Late effective start of NPIs.}
In half of the examined regions, we estimate that the effective start of NPIs $\hat{\tau}$ is significantly later than the official date $\tau^*$.

In Italy, the first case was officially confirmed on Feb~21. School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
That is, the first and last official NPI dates are Mar~5 and Mar~11.
However, we estimate the effective date $\hat{\tau}$ at Mar~17 ($\pm$2.99 days 95\% CI ; \autoref{fig:late}).

@@

In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score: 2--1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~24 ($\pm$2.96 days 95 \%CI), rather than Mar~14 (\autoref{fig:late}).

Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ on Mar~24 ($\pm$2.29 days 95\% CI, \autoref{fig:late}). This is about a week later then the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.

Interestingly, the effect of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
This may suggest correlations between effective starts of NPIs due to global or international events.

@@

We also estimate early effective start of NPIs in Sweden and Denmark.
However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and the fit of the model with free $\tau$ wasn't better than the fit of the model with $\tau$ fixed at the official date (\autoref{table:WAIC}).
In Sweden, the data do not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
We suspect that these missing and additional "peaks" increase the uncertainty in our inference.

@@

%%% Exact
\paragraph*{Exact effective start of NPIs.}
We find one case in which the official and effective dates match "\emph{like a Swiss watch}", and the credible interval is narrow: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\pm$2.51 days 95\% CI, see \autoref{fig:early}). It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.

The estimated effective date in Norway is Mar~22, which is close to the official date of Mar~24.
However, the posterior distribution is very wide ($\pm$10.79 days 95\% CI; \autoref{fig:early}): it covers the range between Mar~9, three days before the first NPI, and Mar~29, five days  after the last NPI (see 
\autoref{table:NPI_dates} and \autoref{fig:NPI_dates} for NPI dates).
The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).

@@

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a generation interval of 4-7 days and neglect the late or early population response to the NPI.
This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$). If health officials assumed the dynamics to immediately change at $\tau^*$, they will have expected the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
This would have lead to a significant underestimation, which might have been interpreted by as ineffectiveness of the NPI, leading to further escalations.
However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data (stars).

%%% Fig - Prediction
\begin{figure}[h]
    \centering
	\includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a generation interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:prediction}
\end{figure}

@@

In most investigated countries we find a late effective start of NPIs on the transmission dynamics.
For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
This difference might be explained, in some cases, by low compliance: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.

In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}), although this is only conclusive in Germany (\autoref{fig:tau-summary}).
An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
Adoption of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may be influenced by increased risk perception due to domestic or international COVID\=/19-related reports~\citep{Arthur2020}.
Indeed, the evidence supports a change in infection dynamics (i.e. a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{table:RMSE}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}. %TODO mobility decreased

Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid~\citep{Bar-On2020}.
However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).

As expected, we have found that the evidence supports a model in which transmission rate changes at a specific time point $\tau$ over a model with a constant transmission rate (i.e. free or fixed model in \autoref{table:RMSE} and \autoref{table:WAIC}).
It may be interesting to check if the evidence favors a model with \emph{two} or more change-points, rather than one. 
Multiple such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementations of NPIs~\citep{Flaxman2020}.

@@

@@

%%% Data %%%%
\paragraph*{Data.} 
We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Apr 2020. These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR) at each day $t$.
Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
Where there were multiple sequences of days with zero confirmed cases (e.g. France), we cropped the data to begin with the last sequence so that our analysis focuses on the first sustained outbreak rather than isolated imported cases. 
For official NPI dates see \autoref{table:NPI_dates}.

@@

We defined the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} 
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\Delta t & \sim \mathit{Uniform}(1, 5) \\
\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 1, T-2\Big),
\end{aligned}
\end{equation}
where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
We also tested an uninformative uniform prior, $\mathit{Uniform}(1,T-2)$.
WAIC (\autoref{eq:WAIC}) of a model with this uniform prior was either higher, or lower by less than 2, compared to WAIC of a model with the truncated normal prior.
The uninformative prior resulted in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
We therefore decided to use the more informative truncated normal prior for $\tau$.
Other priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior for $\Delta t$.
Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.

@@

We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$ fitted to data up to Apr~11, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}) up to Apr~24, and plot the predicted dynamics~(\autoref{fig:ppc}).
Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the tightness of the predictions) are good ways to visually compare models.
We also compute the expected posterior root mean squared error of these predictions~(\autoref{table:RMSE}).

@@

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yinon M. Bar-On, Lilach Hadany and Oren Kolodny for discussions and comments.
This work was supported in part by the Israel Science Foundation 3811/19 and 1399/17.
}

>>>>

Here, we estimate the joint posterior distribution of the effective start date of NPIs, $\tau$, and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$. 
We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
This is done under the simplifying assumption that all interventions start at a specific date, despite the reality that the durations between the first and last NPIs were between 4 and 12 days (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant).
All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~24, and these predictions were then compared to the real case data. 
The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark (\autoref{table:WAIC}).
Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide 95\% credible intervals on $\hat{\tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).

We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$ (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
The exceptions are Switzerland and Norway (see below).
The latter also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).

In the following, we describe our findings in more detail.

%%% Late
\paragraph*{Late effective start of NPIs.}
In half of the examined regions, we estimate that the effective start of NPIs $\hat{\tau}$ is later than the official date $\tau^*$.

In Italy, the first case was officially confirmed on Feb~21. School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
That is, the first and last official NPI dates are Mar~5 and Mar~11.
However, we estimate the effective date $\hat{\tau}$ six days after the lockdown, at Mar~17 ($\pm$2.99 days 95\% CI; \autoref{fig:late}).

@@

In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score 2--1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ ten days later, at Mar~24 ($\pm$2.96 days 95 \%CI), rather than Mar~14 (\autoref{fig:late}).

Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ at Mar~24 ($\pm$2.29 days 95\% CI, \autoref{fig:late}).
This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.

Interestingly, the effective start of of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
This may suggest correlations between effective starts of NPIs due to global or international events.

@@

We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free $\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).

In Sweden, the data do not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
We suspect that these missing and additional "peaks" increase the uncertainty in our inference.

The estimated effective date in Norway is Mar~22, which is two days earlier than the official date of Mar~24.
However, the posterior distribution is very wide ($\pm$10.79 days 95\% CI; \autoref{fig:early}): it covers the range between Mar~9, three days before the first NPI, and Mar~29, five days  after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).

@@

%%% Exact
\paragraph*{Exact effective start of NPIs.}
We find one case in which the official and effective dates match "\emph{like a Swiss watch}" and the credible interval is narrow.
Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\pm$2.51 days 95\% CI, see \autoref{fig:early}). It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.

@@

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a generation interval of 4-7 days and neglect the late or early population response to the NPI.
This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
If health officials assume that the dynamics change exactly at $\tau^*$, they would expect the number of cases to be within the red lines (posterior predictions assuming $\tau=\tau^*$).
This would lead to a significant underestimation, which might be interpreted as ineffectiveness of the NPI, leading to further escalations.
However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data (stars).

%%% Fig - Prediction
\begin{figure}[h]
    \centering
	\includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data; line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a generation interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:prediction}
\end{figure}

@@

In most investigated countries we find a late effective start of NPIs.
For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
This difference might be explained, in some cases, by low compliance: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.

In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented, although this is only conclusive in Germany (\autoref{fig:early}).
An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
Adoption of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may be influenced by increased risk perception due to domestic or international COVID\=/19-related reports~\citep{Arthur2020}.
Indeed, the evidence supports a change in infection dynamics (i.e. a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{table:RMSE}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.

@@

Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid\citep{Bar-On2020}.
However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).

As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (\autoref{table:WAIC} and \autoref{table:RMSE}).
It may be interesting to check if the evidence favors a model with \emph{two} or more change-points, rather than one. 
Multiple change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementations of NPIs~\citep{Flaxman2020}.

@@

%%% Data %%%%
\paragraph*{Data.} 
We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Apr 2020. These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR tests) on each day $t$.
Data for Wuhan, China retrieved from \citet{Pei2020}. Data for 11 European countries retrieved from \citet{Flaxman2020}. 
Where there were multiple sequences of days with zero confirmed cases (e.g. France), we cropped the data to begin with the last sequence so that our analysis focuses on the first sustained outbreak rather than isolated imported cases. 
For official NPI dates see \autoref{table:NPI_dates}.

@@

We defined the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} 
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\Delta t & \sim \mathit{Uniform}(1, 5) \\
\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 1, T-2\Big),
\end{aligned}
\end{equation}
where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
We also tested an uninformative uniform prior, $\mathit{Uniform}(1,T-2)$.
WAIC (\autoref{eq:WAIC}) of a model with this uniform prior was either higher, or lower by less than 2, compared to WAIC of a model with the truncated normal prior.
The uninformative prior resulted in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
We therefore decided to use the more informative truncated normal prior for $\tau$.
Other priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior for $\Delta t$.
We also tried to estimate the value of $p_t$ before and after $\tau$, instead of keeping it fixed at $1/9$ and $1/6$. The model with fixed values was supported by the evidence (lower WAIC, see below) in 9 of 12 countries. Moreover, the estimates for Wuhan, China were $1/9$ and $1/6$, as in \citet{Li2020}. \\
Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.

@@

We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$ fitted to data up to Apr~11, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}) up to Apr~24, and plot the predicted dynamics~(\autoref{fig:ppc}).
Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the tightness of the predictions) are good ways to visually compare models.
We also compute the expected posterior RMSE (root mean squared error) of these predictions~(\autoref{table:RMSE}).

@@

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yinon M. Bar-On, Lilach Hadany, and Oren Kolodny for discussions and comments.
This work was supported in part by the Israel Science Foundation (3811/19, 552/19, and 1399/17).
}
